---
title: "Getting Started with evoFE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with evoFE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is evoFE?

**evoFE** (Evolutionary Feature Engineering) uses a genetic algorithm to
automatically discover useful feature transformations for tabular data. Instead
of manually crafting interaction terms, ratios, or binning strategies, you let
evolution explore the space of possible transformations and keep the ones that
improve predictive performance.

The result is an **evo_recipe** — a reusable transformation pipeline that can be
applied to new data at prediction time.

### How it works

1. **Initialisation** — A population of individuals is created. Each individual
   is a "recipe" containing a set of feature transformations (genes).
2. **Evaluation** — Every individual is scored via cross-validated model
   performance (LightGBM or XGBoost).
3. **Selection** — The top 50 % survive to breed.
4. **Breeding** — Survivors are combined (crossover) and randomly altered
   (mutation) to produce the next generation.
5. **Repeat** — The cycle continues until the fitness plateaus or the
   generation budget is exhausted.

## Installation

```r
# Install from local source
devtools::install_local("path/to/evoFE")

# Or install directly from a git repo
# devtools::install_github("your-org/evoFE")
```

## Quick Start — Binary Classification

Let's classify whether a car has an automatic or manual transmission using the
`mtcars` dataset.

```{r binary-classification}
library(evoFE)

data(mtcars)
df <- mtcars
df$am <- as.integer(df$am)   # target: 0 = automatic, 1 = manual

res <- evolve_features(
  data      = df,
  target_col = "am",
  task       = "classification",
  evaluator  = "xgboost",
  generations = 5,
  pop_size    = 8,
  cv_folds    = 3,
  early_stopping_rounds = 3,
  seed        = 42,
  verbose     = TRUE
)
```

The returned `evo_recipe` object contains the best individual (feature recipe),
the fitted model, and the evolution history.

```{r binary-inspect}
# View the winning recipe
cat("Best recipe:", individual_to_recipe_string(res$best_individual), "\n")
cat("Fitness:    ", res$best_individual$fitness, "\n")
```

### Applying the recipe to new data

`predict()` applies the evolved transformations to new data and returns the
engineered feature matrix:

```{r binary-predict-features}
engineered <- predict(res, df[1:5, ])
head(engineered)
```

`predict_model()` goes one step further — it applies the transformations **and**
runs the trained model to produce predictions:

```{r binary-predict-model}
preds <- predict_model(res, df[1:5, ])
preds
```

## Regression

Predict petal length from the iris dataset:

```{r regression}
data(iris)

res_reg <- evolve_features(
  data      = iris[, 1:5],
  target_col = "Petal.Length",
  task       = "regression",
  evaluator  = "xgboost",
  generations = 5,
  pop_size    = 8,
  cv_folds    = 3,
  early_stopping_rounds = 3,
  seed        = 123,
  verbose     = TRUE
)

cat("Best recipe:", individual_to_recipe_string(res_reg$best_individual), "\n")
cat("Fitness (neg RMSE):", res_reg$best_individual$fitness, "\n")
```

```{r regression-predict}
preds_reg <- predict_model(res_reg, iris[1:10, ])
# Compare predictions to actuals
data.frame(
  actual    = iris$Petal.Length[1:10],
  predicted = round(preds_reg, 2)
)
```

## Multiclass Classification

Classify iris species (3 classes). Note `task = "multiclass"`:
  
```{r multiclass}
iris_mc <- iris
iris_mc$Species <- as.character(iris_mc$Species)

res_mc <- evolve_features(
  data      = iris_mc,
  target_col = "Species",
  task       = "multiclass",
  evaluator  = "xgboost",
  generations = 5,
  pop_size    = 8,
  cv_folds    = 3,
  early_stopping_rounds = 3,
  seed        = 99,
  verbose     = TRUE
)

cat("Best recipe:", individual_to_recipe_string(res_mc$best_individual), "\n")
```

For multiclass, `predict_model()` returns a probability matrix — one column per
class:

```{r multiclass-predict}
probs <- predict_model(res_mc, iris_mc[c(1, 51, 101), ])
round(probs, 3)
```

## Transformer Reference

evoFE ships with **29 built-in transformers** that the genetic algorithm can
select from during evolution. The table below groups them by category.

### Arithmetic (numeric → numeric)

| Transformer | Arity | Description |
|:---|:---:|:---|
| `log` | unary | Natural logarithm (safe: `log(abs(x) + 1)`) |
| `sqrt` | unary | Square root (safe: `sqrt(abs(x))`) |
| `reciprocal` | unary | `1 / (x + ε)` |
| `add` | multi | Element-wise sum of 2+ columns |
| `subtract` | binary | `x₁ − x₂` |
| `multiply` | multi | Element-wise product of 2+ columns |
| `divide` | binary | `x₁ / (x₂ + ε)` |
| `normalized_difference` | binary | `(x₁ − x₂) / (x₁ + x₂ + ε)` |
| `log_ratio` | binary | `log((x₁ + ε) / (x₂ + ε))` |

### Group-by aggregations (mixed → numeric)

These combine a **categorical** grouping column with a **numeric** value column.

| Transformer | Description |
|:---|:---|
| `groupby_mean` | Mean of value within each group |
| `groupby_sd` | Standard deviation within each group |
| `groupby_max` / `groupby_min` | Max / min within each group |
| `groupby_ratio` | `value / group_mean` |
| `groupby_zscore` | `(value − group_mean) / group_sd` |

### Encoding & binning

| Transformer | Input → Output | Description |
|:---|:---:|:---|
| `target_encode` | cat → num | Supervised mean-target encoding with leave-one-out |
| `frequency_encode` | cat → num | Proportion of each category in the data |
| `quantile_binning` | num → num | Assign quantile rank (1–5) |
| `log_binning` | num → num | Assign log-scale bin index |
| `quantile_binning_cat` | num → cat | Same as `quantile_binning`, output as factor |
| `log_binning_cat` | num → cat | Same as `log_binning`, output as factor |

### Dimensionality reduction (numeric → numeric)

| Transformer | Description |
|:---|:---|
| `pca` | First principal component of 2+ columns |
| `truncated_svd` | First component from truncated SVD |
| `random_projection` | Random linear combination of 2+ columns |
| `umap` | Low-dimensional UMAP projection |

### Manifold & Graph Learning (numeric → categorical/numeric)

| Transformer | Output | Description |
|:---|:---:|:---|
| `genie` | categorical | Genie robust hierarchical clustering |
| `lumbermark` | categorical | Lumbermark hierarchical clustering |
| `mst_score` | numeric | Minimum Spanning Tree-based anomaly score |
| `deadwood` | numeric | Deadwood anomaly detection (outlier indicators) |

## Hierarchical Features (Gene Chaining)

One of evoFE's powerful capabilities is **hierarchical feature construction**.
After a gene has been evaluated and proven useful, subsequent generations can
build *on top of* its output.

For example:

```
Gen 1: log_ratio(Sepal.Length, Petal.Width)        → tested ✓
Gen 2: divide(Petal.Width, logratio(…))            → chains from tested gene ✓
```

**Important safety rule**: a gene can only chain from outputs that have been
evaluated in a **previous** generation. A brand-new untested gene is never used
as input for another gene in the same individual. This prevents fragile
dependency chains built on unproven transformations.

## Understanding the Output

`evolve_features()` returns an `evo_recipe` S3 object with:

| Field | Description |
|:---|:---|
| `best_individual` | The winning recipe (list of genes, column sets, fitness) |
| `best_model` | The final LightGBM/XGBoost model trained on all data |
| `history` | Full final-generation population (for inspection) |
| `task` | The task type used |
| `evaluator` | The evaluator used |
| `classes` | Class labels (multiclass only) |

### Inspecting the recipe

```{r inspect-recipe}
ind <- res$best_individual

# Human-readable recipe string
cat(individual_to_recipe_string(ind), "\n")

# Number of evolved genes
cat("Evolved genes:", length(ind$genes), "\n")

# Original columns retained
cat("Numeric cols: ", paste(ind$numeric_cols, collapse = ", "), "\n")
cat("Categorical cols:", paste(ind$categorical_cols, collapse = ", "), "\n")

# Individual gene details
for (g in ind$genes) {
  cat(sprintf("  %s(%s) → %s\n",
    g$transformer_name,
    paste(g$input_cols, collapse = ", "),
    g$output_col))
}
```

## Tuning Parameters

### Key parameters for `evolve_features()`

| Parameter | Default | Description |
|:---|:---:|:---|
| `generations` | 10 | Maximum number of evolutionary generations |
| `pop_size` | 10 | Number of individuals per generation |
| `cv_folds` | 3 | Cross-validation folds for fitness evaluation |
| `early_stopping_rounds` | 3 | Stop if no improvement for *n* generations |
| `evaluator` | `"lightgbm"` | Model backend: `"lightgbm"` or `"xgboost"` |
| `dynamic_population` | `TRUE` | Expand population during stagnation |
| `crossover_type` | `"both"` | `"random"`, `"union"`, or `"both"` |
| `threads` | 8 | Parallelism for model training |
| `seed` | `NULL` | RNG seed for reproducibility |

### Practical advice

- **Start small**: `generations = 5, pop_size = 8` is enough to validate the
  pipeline. Scale up once you confirm the setup works.
- **Increase `pop_size`** for wider exploration. Useful when you have many
  columns (> 20) and diverse transformer options.
- **Increase `generations`** for deeper search. Works best when combined with
  `dynamic_population = TRUE` so stagnation triggers population expansion.
- **Use `seed`** for reproducible experiments and benchmarking.
- **`crossover_type = "union"`** tends to produce larger recipes (more features).
  `"random"` keeps recipes leaner.

## Reproducibility

Setting a seed guarantees identical results across runs:

```{r reproducibility}
r1 <- evolve_features(iris[,1:5], "Petal.Length", task = "regression",
                       generations = 3, pop_size = 5, evaluator = "xgboost",
                       seed = 42, verbose = FALSE)
r2 <- evolve_features(iris[,1:5], "Petal.Length", task = "regression",
                       generations = 3, pop_size = 5, evaluator = "xgboost",
                       seed = 42, verbose = FALSE)

identical(r1$best_individual$fitness, r2$best_individual$fitness)
identical(
  individual_to_recipe_string(r1$best_individual),
  individual_to_recipe_string(r2$best_individual)
)
```

## End-to-End Example: Train/Test Split

A realistic workflow with hold-out evaluation:

```{r end-to-end}
data(iris)
set.seed(1)
idx <- sample(nrow(iris), 0.7 * nrow(iris))
train <- iris[idx, ]
test  <- iris[-idx, ]

# Evolve on training data only
recipe <- evolve_features(
  data      = train[, 1:4],               # exclude Species
  target_col = "Petal.Length",
  task       = "regression",
  evaluator  = "xgboost",
  generations = 5,
  pop_size    = 8,
  seed        = 7,
  verbose     = FALSE
)

# Predict on held-out test data
test_preds <- predict_model(recipe, test[, 1:4])

# Evaluate
rmse <- sqrt(mean((test$Petal.Length - test_preds)^2))
cat(sprintf("Test RMSE: %.4f\n", rmse))
cat(sprintf("Recipe:    %s\n", individual_to_recipe_string(recipe$best_individual)))
```

## Session Info

```{r session-info}
sessionInfo()
```
